---
title: "A zero-inflated Poisson model for cases of Malaria in Tartagal"
author: "Paulo Inácio Prado"
date: "`r format(Sys.time(), '%d de %B de %Y')`"
output: 
        rmdformats::readthedown:
        self_contained: true
        thumbnails: true
        lightbox: true
        gallery: false
        highlight: tango
        toc_depth: 4
---



```{r setup, echo=FALSE, warning=FALSE, message=F}
library(knitr); library(zoo); library(xts)
library(ggplot2)#; library(cowplot)
library(dplyr); library(tidyr); library(ggmap);
library(rjags); library(R2jags)

opts_chunk$set(fig.align = 'center', fig.show = 'hold', fig.height = 4,
               warning = FALSE, message = FALSE, error = FALSE, echo=FALSE)
options(formatR.arrow = TRUE,width = 90, cache=TRUE)
```

# Statistical model used

The idea is to fit an inhomogeneous zero-inflated Poisson dynamics in discrete time.
The expected size of the population of infected people is the single state variable,
which has a growth that can change at each time step. The number of infected people at
each time step follows a Poisson distribution where its single parameters is the expected number of cases.
The number of recorded cases is a proportion of the number of infected people.

This model thus describe the number of observed cases as a zero-inflated Poisson variable,
that evolve in time (that is , inhomogeneous in time). Our purpose with this model is
to provide a time series of expected number of cases with minimal assumptions. 
By replacing the original time series of observed cases with this estimated time series can have two main
advantages:

- Zero values are eliminated, as they are replaced by the expected numbers (the estimated time-varying parameter of the Poisson).
- Observation (including detection errors that can explain part of the zeroes) and part of prcess errors are sorted out.


## Model specification:

##### Number of infected people in the population: 

   * E[N(t+1)] =  N(t) exp(r(t))
   * N(t) ~ Poisson(lambda = E[N(t)])

##### Number of cases detected :

   * n(t) ~ Binomial(P(t), N(t))

##### Where:

* E[X] expected values
* r(t) a time-varying rate of change
* P is the proportion of detected cases


## Model script

The model is a discrete time dynamics, fitted by the a bayesian MCMC routine
using the [JAGS](http://www.uvm.edu/~bbeckage/Teaching/DataAnalysis/Manuals/manual.jags.pdf), language with the R packages
[rjags](https://cran.r-project.org/package=rjags) e [R2jags](https://cran.r-project.org/package=R2jags).

The model script in Jags is [here](poisson_model.jag). 


# Monthly data

* Number of Malaria cases in Tartagal, Salta, from `r format(min((ncases$date)), "%m/%Y")` to 
`r format(max((ncases$date)), "%m/%Y")`
* Time series length: `r length(ncases$date)` months
* Total number of cases: `r sum(ncases$cases)`



```{r data reading}
ncases <- read.csv("../data/DataMalariaTartagalCCM.csv")
ncases$date <- as.Date(ncases$date, "%d/%m/%Y")
#summary(ncases)
ncasesz <- zoo(ncases[,-1], ncases[,1])
plot(ncasesz$cases, ylab="Number of cases")
```

## Model settings

```{r monthly load results}
load("fit1.RData")
tmp <- fit1$BUGSoutput$summary
```

The model was fitted in parallel with a separated R script, because it is time-consuming.

* Number of chains: `r fit1$BUGSoutput$n.chains`
* Number of iterations in each chain: `r fit1$BUGSoutput$n.iter`
* Burn-in period (n iterations discarded): `r fit1$BUGSoutput$n.burnin`
* Thinning period: `r fit1$BUGSoutput$n.thin`
* Number of values of the posterior distribution kept in each interaction: `r fit1$BUGSoutput$n.keep`
* Total number of values of posterior kept: `r fit1$BUGSoutput$n.sims`


## Model Convergence

Convergence ok: 

* R-hat statistics of estimated posterior parameters ranged from `r round(min(fit1$BUGSoutput$summary[,"Rhat"]),4)` 
to `r round(max(fit1$BUGSoutput$summary[,"Rhat"]),4)`
* Effective size of samples of posterior distributions ranged from `r min(fit1$BUGSoutput$summary[,"n.eff"])` 
to `r min(fit1$BUGSoutput$summary[,"n.eff"])`


## Estimated detection probability

The posterior mean detection probability is
`r  round(tmp[grep("^p",rownames(tmp)),"mean"], 3)`.

The plot of posterior density of the four chains pooled is above:

```{r posterior p detection}
plot(as.mcmc(fit1)[,"p"])
```


## Observed x predicted number of cases

The plot shows the expected value of the number of cases recorded and
its credible interval (blue) and the observed number of cases (red).

```{r plots}
## Predicted x observed number of cases
totais <- zoo(
    data.frame(obs=ncases$cases, 
               Ncasos=tmp[grep("ypred",rownames(tmp)),"mean"],
               Ncasos.low=tmp[grep("ypred",rownames(tmp)),"2.5%"],
               Ncasos.up=tmp[grep("ypred",rownames(tmp)),"97.5%"]),
    order.by=as.Date(ncases$date))
## N de casos obs e previsto
totais.df <- totais %>%
    fortify() %>%
    mutate(data=as.Date(Index))
totais.df %>%
    ggplot(aes(data, obs)) +
    geom_line(colour="red", lwd=0.8) +
    geom_line(aes(data, Ncasos), colour="blue")+
    geom_ribbon(aes(ymin=Ncasos.low, ymax=Ncasos.up), fill="blue", alpha=0.25) +
    scale_y_continuous(name="N cases") +
    theme_bw()

## Save the predicted data
write.csv(totais.df, file="predicted_cases_poisson_model_monthly.csv", row.names=FALSE)
```

## Files

* [R script](cluster_months.R) (ran in a separate machine in parallel)
* [R binary file](fit1.RData) with the object of the fitted model (class rjags). 
* [Fitted expected number of cases](predicted_cases_poisson_model_monthly.csv): time series of the mean of posterior distributions. Comma-delimited (csv)


# Weekly data

* Same dataset, but aggregated by epidemiological week.
* Time series length: `r length(ncasesW$data)` weeks


```{r Weekly data reading}
ncasesW <- read.csv("../data/SemanalesTartagal/DatosSemanalesVivaxClimaCompletissima.csv", as.is=TRUE)
ncasesW$data <- as.Date(ncasesW$data)
ncasesWz <- zoo(ncasesW[,-1], ncasesW[,1])
plot(ncasesWz$Vivax.cases, ylab="Number of cases")
```


## Model settings

```{r weekly load results}
load("fit2.RData")
tmp2 <- fit2$BUGSoutput$summary
```

The model was fitted in parallel with a separated R script, because it is time-consuming.

* Number of chains: `r fit2$BUGSoutput$n.chains`
* Number of iterations in each chain: `r fit2$BUGSoutput$n.iter`
* Burn-in period (n iterations discarded): `r fit2$BUGSoutput$n.burnin`
* Thinning period: `r fit2$BUGSoutput$n.thin`
* Number of values of the posterior distribution kept in each interaction: `r fit2$BUGSoutput$n.keep`
* Total number of values of posterior kept: `r fit2$BUGSoutput$n.sims`


## Model Convergence

Convergence ok: 

* R-hat statistics of estimated posterior parameters ranged from `r round(min(fit2$BUGSoutput$summary[,"Rhat"]),4)` 
to `r round(max(fit2$BUGSoutput$summary[,"Rhat"]),4)`
* Effective size of samples of posterior distributions ranged from `r min(fit2$BUGSoutput$summary[,"n.eff"])` 
to `r min((fit2$BUGSoutput$summary[,"n.eff"]))`

The posterior mean detection probability is
`r  round(tmp[grep("^p",rownames(tmp)),"mean"], 3)`.

## Estimated detection probability

The posterior mean detection probability is very low:
`r  round(tmp2[grep("^p",rownames(tmp2)),"mean"], 4)`.

Because this detection is constant, I do not think it will affect
the smoothing of predicted number of cases itself, only teh magnitude of the predicted values.
So, I suppose that's ok to use in CCM, right?

The plot of posterior density of the four chains pooled is below.

```{r weekly posterior p detection}
plot(as.mcmc(fit2)[,"p"])

```


## Observed x predicted number of cases

The plot shows the expected value of the number of cases recorded and
its credible interval (blue) and the observed number of cases (red).


```{r Weekly plots}
## Load the results of the Bayseian fit, that I ran in a cluster
load("fit2.RData")
## Predicted x observed number of cases
tmp2 <- fit2$BUGSoutput$summary
totaisW <- zoo(
    data.frame(obs=ncasesW$Vivax.cases, 
               Ncasos=tmp2[grep("ypred",rownames(tmp2)),"mean"],
               Ncasos.low=tmp2[grep("ypred",rownames(tmp2)),"2.5%"],
               Ncasos.up=tmp2[grep("ypred",rownames(tmp2)),"97.5%"],
               NI=tmp2[grep("N",rownames(tmp2)),"mean"],
               NI.low=tmp2[grep("N",rownames(tmp2)),"2.5%"],
               NI.up=tmp2[grep("N",rownames(tmp2)),"97.5%"]),
    order.by=as.Date(ncasesW$data))
## N de casos obs e previsto
totaisW.df <- totaisW %>%
    fortify() %>%
    mutate(data=as.Date(Index))
totaisW.df %>%
    ggplot(aes(data, obs)) +
    geom_line(colour="red") +
    geom_line(aes(data, Ncasos), colour="blue")+
    geom_ribbon(aes(ymin=Ncasos.low, ymax=Ncasos.up), fill="blue", alpha=0.5) +
    scale_y_continuous(name="N cases") +
    theme_bw()

## Total de doentes
## totaisW.df %>%
##     ggplot(aes(data, NI)) +
##     geom_line() +
##     geom_ribbon(aes(ymin=NI.low, ymax=NI.up),alpha=0.25) +
##     scale_y_continuous(name="N Infected") +
##     theme_bw()

## Simulated number of cases
## A sample of size = 500 from posterior
r.ind <- sample(nrow(fit2$BUGSoutput$sims.list$N))
results
for(i in 1:length(r.ind)){
    
    }
## Save the predicted data
write.csv(totaisW.df, file="predicted_cases_poisson_model_weekly.csv", row.names=FALSE)
```

## Files

* [R script](cluster.R) (ran in a separate machine in parallel)
* [R binary file](fit2.RData) with the object of the fitted model (class rjags). 
* [Fitted expected number of cases](predicted_cases_poisson_model_weekly.csv): time series of the mean of posterior distributions. Comma-delimited (csv)

